---
title: "BFALgtseq-test2"
output: html_notebook
---

11 January 2023

This is the second test of the primers for the BFAL.

Reminding myself: I used two separate primer pools, each with 48/96 samples.



## Taking a look at output from GTscore


```{r}
library(tidyverse)
```


The plot we want is the R1 reads where the FWD primer was found vs. the flashed read where the FWD primer was found.


```{r}
# read in the GTscore output

flashed <- read_tsv("../data/test2/flashed_locusSummary.txt") %>%
  select(Locus, `Primer Reads`) %>%
  rename(flashed_reads = `Primer Reads`)

R1 <- read_tsv("../data/test2/R1_unzipped_locusSummary.txt") %>%
  select(Locus, `Primer Reads`) %>%
  rename(R1_reads = `Primer Reads`)
```

Look at those two df combined based on locus
```{r}
combo_df <- R1 %>%
  left_join(., flashed, by = "Locus")

combo_df %>%
  summarise(sum(R1_reads))

combo_df %>%
  summarise(sum(flashed_reads))
```


```{r}
combo_df %>%
  ggplot(aes(x = flashed_reads, y = R1_reads)) +
  geom_point() +
  labs(title = "BFAL flashed vs. unmerged reads") +
  theme_minimal()


```

So the loci that would be concerning are those with high numbers for the unmerged (R1) reads and disproportionately low numbers for the flashed reads in comparison.

```{r}
# create a proportion difference
combo_df %>%
  mutate(prop = flashed_reads/R1_reads) %>%
  ggplot(aes(x = prop)) +
  geom_histogram() +
  labs(x = "proportion of R1 that also merged",
       title = "BFAL proportion of on-target reads") +
  theme_minimal()


```


More generally, what's the distribution of reads across loci?

```{r}
combo_df %>%
  pivot_longer(cols = 2:3, names_to = "read_type", values_to = "reads") %>%
  group_by(read_type) %>%
  summarise(sum(reads)) 

```

```{r}
1329026/2798712
```
~50% total on-target makes sense given the dimer.

```{r}
combo_df %>%
  filter(!R1_reads == 0) %>% # these correspond to the 18 loci that I removed based on the first iteration.
  mutate(prop = flashed_reads/R1_reads) %>%
  #filter(prop < 0.75 & R1_reads > 10000) %>% # how many loci with < 75% on-target?
  mutate(status = ifelse(prop < 0.75 & R1_reads >5000, "drop", "ok")) %>%
  ggplot(aes(x = flashed_reads, y = R1_reads, color = status)) +
  geom_point() +
  labs(title = "BFAL flashed vs. unmerged reads") +
  theme_minimal()
```
Zoom in on the origin to see what's going on.
```{r}
combo_df %>%
  filter(!R1_reads == 0) %>% # these correspond to the 18 loci that I removed based on the first iteration.
  mutate(prop = flashed_reads/R1_reads) %>%
  #filter(prop < 0.75 & R1_reads > 10000) %>% # how many loci with < 75% on-target?
  mutate(status = ifelse(prop < 0.8 & R1_reads >3900 | flashed_reads > 200000, "drop", "ok")) %>%
  ggplot(aes(x = flashed_reads, y = R1_reads, color = status)) +
  geom_point() +
  geom_abline(linetype = "dotted") +
  labs(title = "BFAL flashed vs. unmerged reads") +
  theme_minimal()

ggsave("pdf_outputs/test2_merged_vs_R1.pdf")
```
Maybe that's too stringent?



```{r}
combo_df %>%
  filter(!R1_reads == 0) %>% # these correspond to the 18 loci that I removed based on the first iteration.
  mutate(prop = flashed_reads/R1_reads) %>%
  mutate(status = ifelse(prop < 0.8 & R1_reads >3900, "drop", "ok")) %>%
  filter(flashed_reads < 10000 & R1_reads < 10000) %>%
  ggplot(aes(x = flashed_reads, y = R1_reads, color = status)) +
  geom_point() +
  labs(title = "BFAL flashed vs. unmerged reads, loci <10k reads") +
  theme_minimal()

ggsave("pdf_outputs/test2_merged_v_R1_under10k.pdf")
```

Now zoom in on loci with > 10k reads

```{r}
combo_df %>%
  filter(!R1_reads == 0) %>% # these correspond to the 18 loci that I removed based on the first iteration.
  mutate(prop = flashed_reads/R1_reads) %>%
  #filter(prop < 0.75 & R1_reads > 10000) %>% # how many loci with < 75% on-target?
  mutate(status = ifelse(prop < 0.75 & R1_reads >3900, "drop", "ok")) %>%
  filter(flashed_reads > 10000 & R1_reads > 10000) %>%
  ggplot(aes(x = flashed_reads, y = R1_reads, color = status)) +
  geom_point() +
  labs(title = "BFAL flashed vs. unmerged reads, loci >10k reads") +
  theme_minimal()
```



The final thing I'm curious about is whether the off-target loci that were retained from the first iteration just received too few reads the first time around to adequately assess them?

Which loci am I planning on removing based on gtseq test 2?
```{r}
test2_droppers <- combo_df %>%
  filter(!R1_reads == 0) %>% # these correspond to the 18 loci that I removed based on the first iteration.
  mutate(prop = flashed_reads/R1_reads) %>%
  #filter(prop < 0.75 & R1_reads > 10000) %>% # how many loci with < 75% on-target?
  mutate(status = ifelse(prop < 0.9 & R1_reads >3900 | flashed_reads > 200000, "drop", "ok")) %>%
  filter(status == "drop")

```


```{r}
# how many remain?
keepers <- combo_df %>%
  filter(!R1_reads == 0) %>% # these correspond to the 18 loci that I removed based on the first iteration.
  mutate(prop = flashed_reads/R1_reads) %>%
  #filter(prop < 0.75 & R1_reads > 10000) %>% # how many loci with < 75% on-target?
  mutate(status = ifelse(prop < 0.9 & R1_reads >3900 | flashed_reads > 200000, "drop", "ok")) %>%
  filter(status == "ok")

keepers
```
That leaves a total of 284 remaining.



Quick look at the 42 I'm thinking of dropping from round 1.

```{r}
# read in the GTscore output

test1_flashed <- read_tsv("../data/flashed_GTscore_locusSummary.txt") %>%
  select(Locus, `Primer Reads`) %>%
  rename(flashed_reads = `Primer Reads`)

test1_R1 <- read_tsv("../data/R1_GTscore_locusSummary.txt") %>%
  select(Locus, `Primer Reads`) %>%
  rename(R1_reads = `Primer Reads`)
```

```{r}
test1_df <- test1_R1 %>%
  left_join(., test1_flashed, by = "Locus")

t1_df <- test1_df %>%
  mutate(prop = flashed_reads/R1_reads) %>%
  mutate(prop = ifelse(prop == "NaN", 0, prop))

test2_droppers %>%
  left_join(., t1_df, by = "Locus") %>%
  filter(prop.y > 0.5) # look at whether there were some that looked decent in round 1 that we kept in

```

Take a look at just a couple of them:

```{r}
combo_df %>%
  filter(!R1_reads == 0) %>% # these correspond to the 18 loci that I removed based on the first iteration.
  mutate(prop = flashed_reads/R1_reads) %>%
  #filter(prop < 0.75 & R1_reads > 10000) %>% # how many loci with < 75% on-target?
  mutate(status = ifelse(prop < 0.8 & R1_reads >3900 | flashed_reads > 200000, "drop", "ok")) %>%
  mutate(status = ifelse(Locus == "BFAL_scaffold_592_558793", "question", status)) %>% # so that one is just the way overamper
  ggplot(aes(x = flashed_reads, y = R1_reads, color = status)) +
  geom_point() +
  geom_abline(linetype = "dotted") +
  labs(title = "BFAL flashed vs. unmerged reads") +
  theme_minimal()


```


## Which loci to drop, which to change the concentration?

```{r}
combo_df %>%
  filter(!R1_reads == 0) %>% # these correspond to the 18 loci that I removed based on the first iteration.
  mutate(prop = flashed_reads/R1_reads) %>%
  #filter(prop < 0.75 & R1_reads > 10000) %>% # how many loci with < 75% on-target?
  mutate(status = ifelse(prop < 0.8 & R1_reads >3900 | flashed_reads > 200000, "drop", "ok")) %>%
  filter(status != "drop") %>%
  ggplot(aes(x = flashed_reads, y = R1_reads, color = status)) +
  geom_point() +
  scale_color_manual(values = "turquoise") +
  geom_abline(linetype = "dotted") +
  labs(title = "BFAL flashed vs. unmerged reads") +
  theme_minimal()

```
Make a list of the keepers and then change the primer concentration
```{r}
combo_df %>%
  filter(!R1_reads == 0) %>% # these correspond to the 18 loci that I removed based on the first iteration.
  mutate(prop = flashed_reads/R1_reads) %>%
  #filter(prop < 0.75 & R1_reads > 10000) %>% # how many loci with < 75% on-target?
  mutate(status = ifelse(prop < 0.8 & R1_reads >3900 | flashed_reads > 200000, "drop", "ok")) %>%
  filter(status != "drop") %>%
  arrange(desc(flashed_reads)) %>%
  ggplot(aes(x = flashed_reads)) +
  geom_histogram()

```
```{r}
test2_new_primer_multipliers <- keepers %>%
  mutate(multiplier = ifelse(flashed_reads >50000, 0.5, 1)) %>%
    mutate(multiplier = ifelse(flashed_reads >100000, 0.25, multiplier)) %>%
      mutate(multiplier = ifelse(flashed_reads <1000, 2, multiplier))


# change format to match platemaps
test2_new_primer_multipliers$Locus <- gsub("BFAL_", "", test2_new_primer_multipliers$Locus)  
```

Now add on the working stock primer plate info to that

```{r}
# read in primer plate map
platemap <- readxl::read_xlsx("../IDTplateSpecs_BFALgtseq_20221012.xlsx")
```


```{r}
# select the relevant variables and make the sequence name match the data
platemap1 <- platemap %>%
  select(`Plate Name`, `Well Position`, `Sequence Name`) %>%
  rename(locus = `Sequence Name`) 


tmp <- gsub("BFAL_", "", platemap1$locus)
tmp1 <- gsub("_R", "", tmp)
tmp2 <- gsub("_F", "", tmp1)
  
platemap1$locus <- tmp2

platemap2 <- platemap1 %>% # since the FWD and REV primers are combined in plates, I only need one plate/well entry per primer set
  filter(str_detect(`Plate Name`, "R"))

```

Assume we're making 1000 ul of new primer pool
```{r}
tmp1 <- platemap2 %>%
  left_join(., test2_new_primer_multipliers, by = c("locus" = "Locus")) %>%
  mutate(previous_uM = 0.5) %>% # this would correspond to 2.5 ul per primer pair
  mutate(new_uM = previous_uM*multiplier) %>%
  mutate(cherrypick_ul = (new_uM*1000)/200) %>% # working conc is 200 uM and volume of primer pool is 1000 
  mutate(cherrypick_ul = ifelse(is.na(cherrypick_ul), 0, cherrypick_ul)) # make zeros explicit

tmp1 %>%
  select(`Plate Name`, `Well Position`, locus, cherrypick_ul) %>%
  filter(cherrypick_ul >0) %>%
  summarise(sum(cherrypick_ul))

```
That's a little off... but the math should be right.

```{r}
tmp1 %>%
   select(`Plate Name`, `Well Position`, locus, cherrypick_ul) %>%
    arrange(`Plate Name`) %>%
  write_csv("csv_outputs/gtseq_test3_primer_cherrypick.csv")
  
```



