---
title: "09-reference-baseline-mixture-analysis"
output: html_notebook
---

23 February 2023



Implementing the functions for reading in genos from the rds files created by microhaplot.

I used `R-main/01-compile-and-tidy-rds-files.R` to generate an rds file with the genotype info in `data/processed`.

The R function uses a minimum read depth of 10 reads for the first allele and 6 reads for the second allele and a minimum allele balance of 0.4 (this can be modified in `R/microhaplot-genos-funcs.R`).

The VCF file used to generate these rds files is `BFAL_reference_filtered.vcf`, currently on Sedna. I might return to this by merging additional samples into that VCF file, but those would be bycatch - not reference samples - and the benefit to doing so is uncertain.




```{r load-libraries}
library(tidyverse)
library(readxl)
library(stringr)
library(lubridate)
```


```{r}
# read in rds file with genotypes
genos_long <- read_rds("../data/processed/called_genos.rds")

head(genos_long)
```

I need to vet both the loci and the individuals for missing data, etc.

## Locus evaluation

How many loci and how many alleles?
```{r}
# alleles
genos_long %>%
  filter(!is.na(allele)) %>%
  select(locus, allele) %>%
  unique()

# loci
genos_long %>%
  filter(!is.na(allele)) %>%
  select(locus, allele) %>%
  unique() %>%
  group_by(locus) %>%
  tally() %>%
  arrange(desc(n))

  
```
484 alleles across 190 loci with between 1-8 alleles per locus.

Missing data:
```{r}
# missing data across loci
locs_to_toss <- genos_long %>%
  group_by(locus) %>%
  mutate(missingness = ifelse(is.na(allele), 1, 0)) %>%
  summarise(sum(missingness)) %>% # 190 loci x2 = total
  filter(`sum(missingness)`>190) %>% # more than 50% missing data
  select(locus) # drop those loci for now and see how the assignment goes

# just the keepers
genos_locs_filtered <- genos_long %>%
  anti_join(., locs_to_toss)

```
```{r}
# summary of remaining loci
genos_locs_filtered %>%
  filter(!is.na(allele)) %>%
  select(locus, allele) %>%
  unique()

genos_locs_filtered %>%
  filter(!is.na(allele)) %>%
  select(locus, allele) %>%
  unique() %>%
  group_by(locus) %>%
  tally() %>%
  arrange(desc(n))

```
398 alleles in 159 loci with 1-8 alleles per locus.


## COME BACK TO THIS...

First, look at the loci for missingness and >2 haplotypes in an individual [This might be masked by the function that reads in the rds file??]

```{r}
genos_long %>%
  group_by(gtseq_run, id, locus) %>% # there should be no more than 2 alleles for a given indiv/locus
  tally() %>%
  filter(n > 2)

```


## Missing data in individuals

Total number of loci = 159
Total number of alleles = 318

Total number of samples = 1,055

```{r}
inds_to_toss <- genos_locs_filtered %>%
  group_by(gtseq_run, id) %>%
  mutate(missingness = ifelse(is.na(allele), 1, 0)) %>%
  summarise(sum(missingness)) %>%
  arrange(desc(`sum(missingness)`)) %>%
  filter(`sum(missingness)` > 63) # remove samples with >20% missing data

# just the keepers
genos_locs_ind_filtered <- genos_locs_filtered %>%
  anti_join(., inds_to_toss)

```
There's a lot of missing data, unfortunately. We might use a higher threshold for missing data for loci to retain more individuals if possible.

50% missing data threshold = 993 inds to keep.

```{r}
974/1055
```
Take a look at that dataset
```{r}
genos_locs_ind_filtered  %>%
  unite(gtseq_run, id, col = "sample", remove = F) %>%
  ggplot(aes(x = reorder(locus, depth), y = reorder(sample, depth), fill = log10(depth))) +
  geom_tile()

```

## self-assignment 

Doing a sanity check with the reference baseline

```{r}
# first make integers of the alleles
alle_idxs <- genos_locs_ind_filtered %>% 
  dplyr::select(gtseq_run, id, locus, gene_copy, allele) %>%
  group_by(locus) %>%
  mutate(alleidx = as.integer(factor(allele, levels = unique(allele)))) %>%
  ungroup() %>%
  arrange(gtseq_run, id, locus, alleidx) # rubias can handle NA's, so no need to change them to 0's

```


Add population information from metadata:
```{r}
# metadata for reference samples
meta1 <- readxl::read_xlsx("../data/BFAL_WGS_01_finalPlateMap.xlsx", sheet = "combinedPlate1And2SamplePops")

samplelist <- read_csv("../data/gtseq5_samplelist.csv") %>%
  mutate(gtseq_run = "gtseq5") %>%
  rename(id = sample)

metadata <- samplelist %>%
  left_join(., meta1, by = c("Sample_ID" = "ind_ID")) %>%
  mutate(population = ifelse(!is.na(pop), pop, population)) %>%
  mutate(population = ifelse(is.na(population), "bycatch", population)) %>%
  select(-pop)

# just reference samples
ref_samples <- metadata %>%
  filter(!population %in% c("bycatch","NTC")) %>%
  select(Sample_ID, gtseq_run, id, population) %>%
  left_join(., genos_locs_ind_filtered, by = c("gtseq_run", "id"))
  
ref_pops <- ref_samples %>%
  select(gtseq_run, id, Sample_ID, population) %>%
  unique()
```

```{r}
alle_idxs %>%
  filter(gtseq_run == "gtseq5") %>%
  select(id) %>%
  unique() %>%
  inner_join(., ref_pops) %>%
  left_join(., alle_idxs)

```



```{r}
# format for rubias
reference <- alle_idxs %>%
  inner_join(., ref_pops) %>%
  select(-allele, -gtseq_run, -id) %>%
  select(population, Sample_ID, everything()) %>%
  rename(collection = population, indiv = Sample_ID)

# make two-col format
ref_two_col <- reference %>%
  unite("loc", 3:4, sep = ".") %>%
  pivot_wider(names_from = loc, values_from = alleidx) %>%
  mutate(repunit = collection) %>%
  mutate(sample_type = "reference") %>%
  select(sample_type, repunit, collection, everything())

```


```{r}
# self-assignment
baseline_assign <- rubias::self_assign(ref_two_col, gen_start_col = 5)

```

```{r}
top_assign <- baseline_assign %>%
  group_by(indiv) %>%
  slice_max(., order_by = scaled_likelihood) 

top_assign %>% # top assignment for each sample
  ggplot(aes(x = scaled_likelihood)) +
  geom_histogram()

```
Looks like below 90% would be a reasonable cut-off?

```{r}
top_assign %>%
  filter(repunit != inferred_collection, 
         scaled_likelihood > 0.9) %>%
  ggplot(aes(x = scaled_likelihood)) +
  geom_histogram()

```
10 mis-assignments, but only 4 with a scaled-likelihood > 0.9 and all of those are within the Hawaiian colonies.

Pretty good!

So overall:

```{r}
top_assign %>%
  filter(scaled_likelihood > 0.9 &
         collection == inferred_collection)
```
129 samples in the reference baseline (10 dropped out bec of missing data).

90% of samples correctly assigned at 90% scaled likelihood.
```{r}
117/129
```
If we only look at samples that were assigned at > 90% threshold:
```{r}
top_assign %>%
  filter(scaled_likelihood > 0.9)
```
```{r}
117/121
```
Which would be 97% accurate assignment.

Quick look at z-scores:

```{r}
top_assign %>%
  ggplot(aes(z_score)) +
  geom_histogram()


```
Woah. There's an outlier.

```{r}
top_assign %>%
  filter(z_score < -3)

```
Maybe a different species accidentally? I can double-check the sample number with metadata that Jessie sent. Other than that, things are looking good to proceed with mixture assignment.


## Mixture bycatch assignment


```{r}
# get the format right
mix_two_col <- alle_idxs %>%
  anti_join(., ref_pops) %>%
  unite(gtseq_run, id, col = "indiv", sep = "_") %>%
  select(-allele) %>%
  unite("loc", 2:3, sep = ".") %>%
  pivot_wider(names_from = loc, values_from = alleidx) %>%
  mutate(repunit = NA) %>%
  mutate(sample_type = "mixture") %>%
  mutate(collection = "bycatch") %>%
  select(sample_type, repunit, collection, everything())

mix_two_col$repunit <- as.character(mix_two_col$repunit)
head(mix_two_col)

```

```{r}
# for whatever reason, rubias is being finicky about the format and it's easiest to do this
# to make sure both df are consistent
tmp <- bind_rows(ref_two_col, mix_two_col)

mix <- tmp %>%
  filter(sample_type == "mixture")

ref <- tmp %>%
  filter(sample_type == "reference")
```


```{r}
# mixture analysis
bycatch_assign <- rubias::infer_mixture(reference = ref, mixture = mix, gen_start_col = 5)

```

Bycatch results for mixture
```{r}
mix_assigned <- bycatch_assign$indiv_posteriors %>%
  group_by(indiv) %>%
  slice_max(., order_by = PofZ)


# distribution of PofZ for top assignments
mix_assigned %>%
  ggplot(aes(x = PofZ)) +
  geom_histogram()

```
845 bycatch samples, most of which are assigned at high probability.


```{r}
mix_assigned %>%
  filter(PofZ > 0.9)

```

Before going deeper into the assignment results, let's just confirm the loci we're using are okay.


Output genotypes/data for evaluating the loci in `10-locus-summary-and-eval.Rmd`

```{r}
# save outputs
tmp %>%
  write_csv("csv_outputs/two_column_dataset.csv")

genos_locs_ind_filtered %>%
  write_rds("../data/processed/genotypes_loc_ind_filtered.rds")

alle_idxs %>%
  write_rds("../data/processed/alle_idxs.rds")

reference %>%
  write_rds("../data/processed/reference_genos.rds")
```

