---
title: "metadata"
output: html_notebook
---

27 February 2023


Organizing the metadata - linking the MiSeq run sample sheet to the Oikonos metadata and then the genotypes.


```{r}
library(tidyverse)
library(data.table)

# read in metadata from Oikonos
meta <- read_csv("../data/metadata/Oikonos_BFAL_Metadata_12.21.2022.csv") %>%
  select(-c(20:27))

```



```{r}
#create a list of the files from your target directory
file_list <- list.files(path="../data/samplesheets/")

#initiate a blank data frame, each iteration of the loop will append the data from the given file to this variable
samplesheets <- data.frame()

# read in sample sheets
for (i in 1:length(file_list)){
  temp_data <- read_csv(paste0("../data/samplesheets/",file_list[i]), skip = 19) #read in files using the fread function from the data.table package
  samplesheets <- rbindlist(list(samplesheets, temp_data), use.names = T) #for each iteration, bind the new data to the building dataset
}

samplesheets1 <- samplesheets %>%
  rename(id = s_number, gtseq_run = Description) 



samplesheets1$Sample_ID <- gsub("_", "-", samplesheets1$Sample_ID)
samplesheets1$Sample_ID <- gsub("banded", "", samplesheets1$Sample_ID)

samplesheets2 <- samplesheets1 %>%  # one ID to fix based on Claire's description of mislabeling
  mutate(Sample_ID = ifelse(Sample_ID == "BFAL08-1144", "BFAL08-1844", Sample_ID))
  
```

Importantly, I need the plate/gtseq run and the s_number as unique identifiers.


```{r}
# merge sample id with genotype data
# tmp_genos <- genos %>%
#   left_join(., samplesheets1, by = c("gtseq_run", "id")) %>%
#   select(Sample_ID, gtseq_run, id, locus, gene_copy, allele, depth, allele.balance)
# 
# tmp_genos$Sample_ID <- gsub("BFAL", "", tmp_genos$Sample_ID)

```



But what I actually need is a link between the metadata and the samplesheet
```{r}
samplesheets2$Sample_ID <- gsub("BFAL", "", samplesheets2$Sample_ID)

meta_w_samplesheet <- meta %>%
  left_join(., samplesheets2, by = c("MWVCRC#" = "Sample_ID"))

meta_w_samplesheet %>%
  write_csv("../data/processed/metadata_w_samplesheet.csv")

```


```{r}
# Are we missing any samples from the metadata - things that we didn't genotype, e.g.
meta_w_samplesheet %>%
  filter(is.na(gtseq_run))

```
We'll have to double-check that those made it here and are in Claire's office.


Are there genotypes that have no metadata accounted for?

```{r}
# samplesheet vs. metadata
samplesheets2 %>%
  anti_join(., meta, by = c("Sample_ID" = "MWVCRC#")) %>%
  filter(Sample_Plate != "BFAL002", Sample_Plate != "RockfishPlate13", !str_detect(Sample_ID, "NTC"), !str_detect(Sample_ID, "blank")) %>%
  write_csv("csv_outputs/genotypes_wo_metadata.csv")


meta %>%
  filter(str_detect(`MWVCRC#`, "banded"))

```

