---
title: "metadata"
output: html_notebook
---

27 February 2023


Organizing the metadata - linking the MiSeq run sample sheet to the Oikonos metadata and then the genotypes.


```{r}
library(tidyverse)
library(data.table)

# read in metadata from Oikonos
meta <- read_csv("../data/metadata/Oikonos_BFAL_Metadata_12.21.2022.csv") %>%
  select(-c(20:27))

# create some consistent designations for banded locations
banded_meta_locs <- meta %>%
  #filter(`BandY/N` == "TRUE") %>%
  #select(Location, `Loc-Abbr`) %>%
  #unique() %>%
   mutate(Location = ifelse(Location == "French Frigate Shoals (Tern Island)", "Tern Island, French Frigate Shoals, HI", Location)) %>%
    mutate(Location = ifelse(Location == "Laysan Island", "Laysan Island, NWHI, Hawaii", Location)) %>%
      mutate(Location = ifelse(Location == "Kure Attol (Green Island)", "Green Island, Kure Atoll, Honolulu County, Hawaii", Location)) %>%
       mutate(`Loc-Abbr` = ifelse(Location == "Tern Island, French Frigate Shoals, HI", "NWHI, FFS", `Loc-Abbr`)) %>%
        mutate(`Loc-Abbr` = ifelse(Location == "Laysan Island, NWHI, Hawaii", "NWHI, Laysan", `Loc-Abbr`)) %>%
          mutate(`Loc-Abbr` = ifelse(Location == "French Frigate Shoals", "NWHI, FFS", `Loc-Abbr`)) %>%
            mutate(`Loc-Abbr` = ifelse(Location == "Green Island, Kure Atoll, Honolulu County, Hawaii", "NWHI, Kure", `Loc-Abbr`)) %>%
  unique()

# banded_meta_locs %>%
#   group_by(CollYear) %>%
#   tally() 

```



Now organize the sample info associated with the birds we genotyped at ABL
```{r}
#create a list of the files from your target directory
file_list <- list.files(path="../data/samplesheets/")

#initiate a blank data frame, each iteration of the loop will append the data from the given file to this variable
samplesheets <- data.frame()

# read in sample sheets
for (i in 1:length(file_list)){
  temp_data <- read_csv(paste0("../data/samplesheets/",file_list[i]), skip = 19) #read in files using the fread function from the data.table package
  samplesheets <- rbindlist(list(samplesheets, temp_data), use.names = T) #for each iteration, bind the new data to the building dataset
}

samplesheets1 <- samplesheets %>%
  rename(id = s_number, gtseq_run = Description) 



samplesheets1$Sample_ID <- gsub("_", "-", samplesheets1$Sample_ID)
samplesheets1$Sample_ID <- gsub("banded", "", samplesheets1$Sample_ID)

samplesheets2 <- samplesheets1 %>%  # one ID to fix based on Claire's description of mislabeling
  mutate(Sample_ID = ifelse(Sample_ID == "BFAL08-1144", "BFAL08-1844", Sample_ID))
  
```

How many of those IDs are unique?
```{r}
# just the BFAL samples 
ABL_samples_geno <- samplesheets2 %>%
  filter(!str_detect(Sample_ID, "NTC"),
         !str_detect(Sample_ID, "DY"),
         !str_detect(Sample_ID, "blank"),
         !str_detect(Sample_ID, "KURE"),
         !str_detect(Sample_ID, "PCMB"),
         !str_detect(Sample_ID, "TOR")) %>%
  select(Sample_ID) %>%
  unique()

ABL_samples_geno$Sample_ID <- gsub("BFAL", "", ABL_samples_geno$Sample_ID)
```
963 unique BFAL sample IDs

```{r}
869-858
```
```{r}
jessie_meta <- banded_meta_locs %>%
  rename(Sample_ID = `MWVCRC#`)

# which samples have genotypes but no apparent metadata?
extra_genos <- ABL_samples_geno %>%
  anti_join(., jessie_meta) %>%
  filter(!Sample_ID %in% c("11-0293-reference", "19-0164a", "19-0164b"))

extra_genos %>%
  left_join(jessie_meta)
```
11-0293-reference	is almost certainly genotyped twice (1x on the ascertainment plate and again)

```{r}
jessie_meta %>%
  filter(Sample_ID == "19-0164")

```
#### Samples with metadata but no genotypes?

```{r}
jessie_meta %>%
  anti_join(., ABL_samples_geno)

```
"11-0293-reference", "19-0164a", "19-0164b"

So the only missing samples are 18-0517 and 21-0287, which is consistent with what Claire found when she dug through the envelopes in her office (see email).



#### Duplicate samples

```{r}
pot_dups <- samplesheets2 %>%
  filter(!str_detect(Sample_ID, "NTC"),
         !str_detect(Sample_ID, "DY"),
         !str_detect(Sample_ID, "blank"),
         !str_detect(Sample_ID, "KURE"),
         !str_detect(Sample_ID, "PCMB"),
         !str_detect(Sample_ID, "TOR")) %>%
  select(Sample_ID, Sample_Plate) %>%
  group_by(Sample_ID) %>%
  tally() %>%
  filter(n > 1)

duplicate_tissue <- pot_dups %>%
  left_join(., samplesheets2) %>%
  arrange(Sample_ID) %>%
  filter(!Sample_Plate %in% c("BFAL001", "BFAL_bycatch_plate2")) %>%
  group_by(Sample_ID) %>%
  tally() %>%
  filter(n > 1) %>%
  select(Sample_ID)

duplicate_tissue$Sample_ID <- gsub("BFAL", "", duplicate_tissue$Sample_ID)

duplicate_tissue %>%
  left_join(., jessie_meta)
```


A few important notes:
there were samples with modified id numbers in the sample sheet for a couple of purposes. 
Those are:

11-0293-reference
19-0164a

I need to modify those in the sample sheet info so they can be treated as identical samples/duplicates and one of the duplicates can be removed.



## Clean up and merge the metadata for portability to other Rmd docs.

```{r}
samplesheets_fixed <- samplesheets2 %>%
  mutate(Sample_ID = ifelse(Sample_ID == "BFAL11-0293-reference", "BFAL11-0293", Sample_ID)) %>%
  mutate(Sample_ID = ifelse(Sample_ID %in% c("BFAL19-0164a", "BFAL19-0164b"), "BFAL19-0164", Sample_ID))


# metadata for reference samples
jessie_samples_sheet <- samplesheets_fixed %>% # just the ones that were from Jessie
  filter(!str_detect(Sample_ID, "NTC"),
         !str_detect(Sample_ID, "DY"),
         !str_detect(Sample_ID, "blank"),
         !str_detect(Sample_ID, "KURE"),
         !str_detect(Sample_ID, "PCMB"),
         !str_detect(Sample_ID, "TOR"))

# correct format for merging with metadata
jessie_samples_sheet$Sample_ID <- gsub("BFAL", "", jessie_samples_sheet$Sample_ID)

# make changes based on jessie's updated metadata info
jessie_samples_sheet %>%
  filter(!Sample_ID %in% c("11-0626",
                          "11-0628",
                          "16-0412",
                          "16-0413",
                          "19-0314",
                          "12-0369",
                          "17-0268")) %>%
  mutate(Sample_ID = ifelse(Sample_ID == "20-0317", "20-0316", Sample_ID))

# add metadata to that based on sample id
new_meta <- read_xlsx("../data/metadata/Oikonos_NOAA_BFAL_Metadata_updated 3.15.23.xlsx")

Oikonos_samples_meta <- jessie_samples_sheet %>%
  left_join(., new_meta, by = c("Sample_ID" = "MWVCRC#"))


```


Now dealing with the metadata for reference samples not supplied by Jessie:

```{r}
non_jessie_reference_samplesheet <- samplesheets_fixed %>%
  filter(str_detect(Sample_ID, "KURE") |
         str_detect(Sample_ID, "PCMB") |
         str_detect(Sample_ID, "TOR"))

# combine those with the Oikonos samples
full_metadata_samplesheets <- Oikonos_samples_meta %>%
  full_join(., non_jessie_reference_samplesheet) %>%
   mutate(Location = ifelse(str_detect(Sample_ID, "KURE"), "Kure Atoll, Honolulu County, Hawaii", Location)) %>%
     mutate(Location = ifelse(str_detect(Sample_ID, "PCMB"), "Midway Atoll, Hawaii", Location)) %>%
       mutate(Location = ifelse(str_detect(Sample_ID, "TOR"), "Torishima, Japan", Location)) %>%
         mutate(`Loc-Abbr` = ifelse(Location == "Kure Atoll, Honolulu County, Hawaii", "NWHI, Kure", `Loc-Abbr`)) %>%
           mutate(`Loc-Abbr` = ifelse(Location == "Midway Atoll, Hawaii", "NWHI, Midway", `Loc-Abbr`)) %>%
            mutate(`Loc-Abbr` = ifelse(Location == "Torishima, Japan", "Japan, Torishima", `Loc-Abbr`))


full_metadata_samplesheets %>%
  write_csv("../data/processed/full_metadata_w_samplesheet_20230316.csv")

```


Just to confirm everything looks right:
```{r}
full_metadata_samplesheets %>%
  select(Sample_ID) %>%
  unique()

  # tally() %>%
  # filter(n > 1)
```

961 unique samples

These are the samples we want to connect with the genotypes.

```{r}
# read in rds file with genotypes
genos_long <- read_rds("../data/processed/called_genos_na_explicit.rds")
  
genos_w_metadata <- full_metadata_samplesheets %>%
  left_join(., genos_long) %>%
  select(-Sample_Well, -I7_Index_ID, -I5_Index_ID, -index, -index2, -Sample_Project, -`BFALGenetics?`, -VerifySpecies, -BFALColonyGenetics)
  
```


```{r}
# it would be great to add on if a sample was included in the reference
genos_w_metadata

```

```{r}
# from the MiSeq sample sheet
ref_samplelist <- read_csv("../data/gtseq5_samplelist.csv") %>%
  mutate(gtseq_run = "gtseq5") %>%
  rename(id = sample)

# metadata for reference samples used for lcWGS
ref_pops <- readxl::read_xlsx("../data/BFAL_WGS_01_finalPlateMap.xlsx", sheet = "combinedPlate1And2SamplePops")

# combine samplesheet and reference sample info
reference_samplesheet <- ref_samplelist %>%
  left_join(., ref_pops, by = c("Sample_ID" = "ind_ID")) %>%
  mutate(population = ifelse(!is.na(pop), pop, population)) %>%
  mutate(population = ifelse(is.na(population), "bycatch", population)) %>%
  select(-pop, -Sample_ID, -Sample_Plate, -Sample_Well) %>%
  filter(!population %in% c("NTC", "bycatch")) %>%
   mutate(Location = ifelse(population == "Tern", "Tern Island, French Frigate Shoals, HI", NA)) %>%
    mutate(Location = ifelse(population == "Laysan", "Laysan Island, NWHI, Hawaii", Location)) %>%
      mutate(Location = ifelse(population == "Kure", "Kure Atoll, Honolulu County, Hawaii", Location)) %>%
       mutate(Location = ifelse(population == "Torishima", "Torishima Island, Japan", Location)) %>%
        mutate(Location = ifelse(population == "Whale-Skate", "Whale-Skate Island, French Frigate Shoals, HI", Location)) %>%
          mutate(Location = ifelse(population == "Lehua", "Lehua Island, Kauai County, Hawaii", Location)) %>%
          mutate(Location = ifelse(population == "Midway", "Midway Atoll, NWHI, Hawaii", Location)) %>%
       mutate(`Loc-Abbr` = ifelse(Location == "Tern Island, French Frigate Shoals, HI", "NWHI, FFS", NA)) %>%
        mutate(`Loc-Abbr` = ifelse(Location == "Laysan Island, NWHI, Hawaii", "NWHI, Laysan", `Loc-Abbr`)) %>%
          mutate(`Loc-Abbr` = ifelse(Location == "French Frigate Shoals", "NWHI, FFS", `Loc-Abbr`)) %>%
            mutate(`Loc-Abbr` = ifelse(Location == "Whale-Skate Island, French Frigate Shoals, HI", "NWHI, FFS", `Loc-Abbr`)) %>%
         mutate(`Loc-Abbr` = ifelse(Location == "Midway Atoll, NWHI, Hawaii", "NWHI, Midway", `Loc-Abbr`)) %>%
         mutate(`Loc-Abbr` = ifelse(Location == "Lehua Island, Kauai County, Hawaii", "HI, Lehua", `Loc-Abbr`)) %>%
  mutate(`Loc-Abbr` = ifelse(Location == "Kure Atoll, Honolulu County, Hawaii", "NWHI, Kure", `Loc-Abbr`)) %>%
    mutate(`Loc-Abbr` = ifelse(Location == "Torishima Island, Japan", "Japan, Torishima", `Loc-Abbr`)) %>%
  mutate(reference = "TRUE") %>%
  arrange(id)
  

# get the genos and other metadata info for those samples
ref_genos_w_meta <- reference_samplesheet %>%
  left_join(., genos_w_metadata,  by = c("gtseq_run", "id")) %>%
  select(-Location.y, -`Loc-Abbr.y`) %>%
  rename(Location = Location.x, `Loc-Abbr` = `Loc-Abbr.x`) %>%
  select(Sample_ID, population, reference, Sample_Plate, gtseq_run, id, SBNDID, FisheriesName, GenCollArea, `BandY/N`, MinAgeCollection, Location, `Loc-Abbr`, locus, allele, depth, allele.balance)

# merge that with the genos based on gtseq run and id
full_genos_and_meta <- genos_w_metadata %>%
  anti_join(., reference_samplesheet, by = c("gtseq_run", "id")) %>%
  mutate(population = Location, reference = NA) %>%
  select(Sample_ID, population, reference, everything()) %>%
  bind_rows(ref_genos_w_meta)


full_genos_and_meta %>%
  filter(!is.na(reference)) %>%
  select(population, Sample_ID) %>%
  unique()

full_genos_and_meta %>%
  write_rds("../data/processed/genos_w_metadata_samplesheets.rds")

```








## New metadata from Jessie

We learned a few things from Jessie, including, some of the birds with no metadata are not BFAL and should be excluded:

11-0626 - not a BFAL, remove from study
11-0628 - not a BFAL, remove from study
20-0317 - Is this supposed to be 20-0316?; 20-0317 isn't in my records and I shouldn't have had any samples from that animal (which was a harbor seal)
16-0412 - not a BFAL, remove from study
16-0413 - not a BFAL, remove from study
19-0093 - confirmed BFAL (in updated metadata, also banded from Kure if we need to boost baseline samples)
19-0266 - confirmed BFAL (in updated metadata)
19-0314 - not a BFAL, remove from study
13-0999 - confirmed BFAL (in updated metadata)
12-0369 - not a BFAL, remove from study
17-0268 - not a BFAL, remove from study


```{r}
new_meta <- read_xlsx("../data/metadata/Oikonos_NOAA_BFAL_Metadata_updated 3.15.23.xlsx")


```
849 samples in the bycatch, including the banded birds that were part of the baseline.

```{r}
samplesheets_fixed$Sample_ID <- gsub("BFAL", "", samplesheets_fixed$Sample_ID)


samplesheets_keepers <- samplesheets_fixed %>%
  filter(!Sample_ID %in% c("11-0626",
                          "11-0628",
                          "16-0412",
                          "16-0413",
                          "19-0314",
                          "12-0369",
                          "17-0268")) %>%
  mutate(Sample_ID = ifelse(Sample_ID == "20-0317", "20-0316", Sample_ID))

```


```{r}
bycatch_meta_samplesheets <- new_meta %>%
  left_join(., samplesheets_keepers, by = c("MWVCRC#" = "Sample_ID"))

samplesheets_keepers %>%
  left_join(., full_metadata_samplesheets) %>%
  anti_join(., bycatch_meta_samplesheets, by = c("Sample_ID" = "MWVCRC#"))

```

















## old stuff below here






Importantly, I need the plate/gtseq run and the s_number as unique identifiers.


```{r}
# merge sample id with genotype data
# tmp_genos <- genos %>%
#   left_join(., samplesheets1, by = c("gtseq_run", "id")) %>%
#   select(Sample_ID, gtseq_run, id, locus, gene_copy, allele, depth, allele.balance)
# 
# tmp_genos$Sample_ID <- gsub("BFAL", "", tmp_genos$Sample_ID)

```



But what I actually need is a link between the metadata and the samplesheet
```{r}
samplesheets2$Sample_ID <- gsub("BFAL", "", samplesheets2$Sample_ID)

meta_w_samplesheet <- banded_meta_locs %>%
  left_join(., samplesheets2, by = c("MWVCRC#" = "Sample_ID"))

meta_w_samplesheet %>%
  write_csv("../data/processed/metadata_w_samplesheet.csv")

```


```{r}
# Are we missing any samples from the metadata - things that we didn't genotype, e.g.
meta_w_samplesheet %>%
  filter(is.na(gtseq_run))

```
3 samples that we don't have a record of genotyping. We can follow up on that later.



Are there genotypes that have no metadata accounted for?

```{r}
# samplesheet vs. metadata
samplesheets2 %>%
  anti_join(., meta, by = c("Sample_ID" = "MWVCRC#")) %>%
  filter(Sample_Plate != "BFAL002", Sample_Plate != "RockfishPlate13", !str_detect(Sample_ID, "NTC"), !str_detect(Sample_ID, "blank")) %>%
  write_csv("csv_outputs/genotypes_wo_metadata.csv")


meta %>%
  filter(str_detect(`MWVCRC#`, "banded"))

```
Potentially 13 (because 11-0293-reference is just a mismatch with the modification to the ID).
