---
title: "01-metadata"
author: "Diana Baetscher"
date: "2023-03-30"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


30 March 2023


Cleaning up final set of metadata before R&R final report and passing off this repo to Jessie.


## Metadata for bycatch and reference samples


```{r}
library(tidyverse)
library(data.table)
library(readxl)

```

Metadata from Jessie for the bycatch and banded bycatch birds
```{r fix-up-the-new-metadata}
# while I'm at it, let's clean up location info for this
# add metadata to that based on sample id
new_meta <- read_xlsx("../data/metadata/Oikonos_NOAA_BFAL_Metadata_updated 3.15.23.xlsx")


# all possible location info syntax
meta_fixed_locs <- new_meta %>%
  #select(Location, `Loc-Abbr`) %>%
  #unique() %>%
   mutate(Location = ifelse(Location == "French Frigate Shoals (Tern Island)", "Tern Island, French Frigate Shoals, HI", Location)) %>%
    mutate(Location = ifelse(Location == "Laysan Island", "Laysan Island, Honolulu County, Hawaii", Location)) %>%
      mutate(Location = ifelse(Location == "Kure Attol (Green Island)", "Green Island, Kure Atoll, Honolulu County, Hawaii", Location)) %>%
       mutate(`Loc-Abbr` = ifelse(Location == "Tern Island, French Frigate Shoals, HI", "NWHI, FFS", `Loc-Abbr`)) %>%
        mutate(`Loc-Abbr` = ifelse(Location == "Laysan Island, Honolulu County, Hawaii", "NWHI, Laysan", `Loc-Abbr`)) %>%
          mutate(`Loc-Abbr` = ifelse(Location == "French Frigate Shoals", "NWHI, FFS", `Loc-Abbr`)) %>%
            mutate(`Loc-Abbr` = ifelse(Location == "Green Island, Kure Atoll, Honolulu County, Hawaii", "NWHI, Kure", `Loc-Abbr`)) %>%
   unique() #%>% # there are three metadata entries that are exact duplicates. Remove those here
  # arrange(Location)

```
846 bycatch, some of which were banded and used as reference birds



Metadata from the reference birds (non-banded)

```{r}
# from the MiSeq sample sheet
ref_samplelist <- read_csv("../data/gtseq5_samplelist.csv") %>%
  mutate(gtseq_run = "gtseq5") %>%
  rename(id = sample)

# metadata for reference samples used for lcWGS
ref_pops <- readxl::read_xlsx("../data/BFAL_WGS_01_finalPlateMap.xlsx", sheet = "combinedPlate1And2SamplePops") %>%
  mutate(reference = TRUE) %>%
  rename(sampleID = ind_ID)
  


reference_meta_wo_banded <- ref_pops %>%
  left_join(., ref_samplelist, by = c("sampleID" = "Sample_ID")) %>%
  filter(str_detect(sampleID, "BFAL-")) %>% # just the reference samples that came from non-Jessie sources (not banded birds)
  rename(Location = population) %>%
  mutate(Location = ifelse(Location == "Kure", "Kure Atoll, Honolulu County, Hawaii", Location)) %>%
   mutate(Location = ifelse(Location == "Midway", "Midway Atoll, NWHI, HI", Location)) %>%
      mutate(Location = ifelse(Location == "Torishima", "Torishima Island, Japan", Location)) %>%
       mutate(`Loc-Abbr` = ifelse(Location == "Midway Atoll, NWHI, HI", "NWHI, Midway", NA)) %>%
        mutate(`Loc-Abbr` = ifelse(Location == "Torishima Island, Japan", "Japan, Torishima", `Loc-Abbr`)) %>%
            mutate(`Loc-Abbr` = ifelse(Location == "Kure Atoll, Honolulu County, Hawaii", "NWHI, Kure", `Loc-Abbr`)) 
  
reference_meta_wo_banded
```
92 reference samples without banded birds.



## Samplesheets

Now organize the sample info associated with the birds we genotyped at ABL
```{r}
#create a list of the files from your target directory
file_list <- list.files(path="../data/samplesheets/")

#initiate a blank data frame, each iteration of the loop will append the data from the given file to this variable
samplesheets <- data.frame()

# read in sample sheets
for (i in 1:length(file_list)){
  temp_data <- read_csv(paste0("../data/samplesheets/",file_list[i]), skip = 19) #read in files using the fread function from the data.table package
  samplesheets <- rbindlist(list(samplesheets, temp_data), use.names = T) #for each iteration, bind the new data to the building dataset
}

samplesheets1 <- samplesheets %>%
  rename(id = s_number, gtseq_run = Description) 



samplesheets1$Sample_ID <- gsub("_", "-", samplesheets1$Sample_ID)
samplesheets1$Sample_ID <- gsub("banded", "", samplesheets1$Sample_ID)

samplesheets2 <- samplesheets1 %>%  # one ID to fix based on Claire's description of mislabeling
  mutate(Sample_ID = ifelse(Sample_ID == "BFAL08-1144", "BFAL08-1844", Sample_ID)) %>%
  mutate(Sample_ID = ifelse(Sample_ID == "BFAL11-0293-reference", "BFAL11-0293", Sample_ID)) %>%
  mutate(Sample_ID = ifelse(Sample_ID %in% c("BFAL19-0164a", "BFAL19-0164b"), "BFAL19-0164", Sample_ID))


# How many of those IDs are unique?

# just the BFAL samples that are part of the bycatch metadata from Jessie
bycatch_samplesheet <- samplesheets2 %>%
  filter(!str_detect(Sample_ID, "NTC"),
         !str_detect(Sample_ID, "DY"),
         !str_detect(Sample_ID, "blank"),
         !str_detect(Sample_ID, "KURE"),
         !str_detect(Sample_ID, "PCMB"),
         !str_detect(Sample_ID, "TOR")) %>%
  rename(sampleID = Sample_ID)

# Add info about which bycatch samples are in the reference panel?
bycatch_samplesheet_data <- bycatch_samplesheet %>%
  left_join(., ref_pops)

# fix the sample IDs to match the metadata
bycatch_samplesheet_data$sampleID <- gsub("BFAL", "", bycatch_samplesheet_data$sampleID)

# these should now match up with Jessie's metadata for bycatch
bycatch_samplesheet_data 

```




```{r combine-bycatch-and-reference-metadata}
 full_metadata_all_samples <- bycatch_samplesheet_data %>%
  full_join(., meta_fixed_locs, by = c("sampleID" = "MWVCRC#")) %>%
  bind_rows(., reference_meta_wo_banded) %>%
  select(-I7_Index_ID, -index, -I5_Index_ID, -index2, -Sample_Project) %>%
  rename(population = pop)

full_metadata_all_samples %>%
  write_rds("../data/processed/metadata_bycatch_and_reference_20230330.rds")

```

Because some samples were genotyped multiple times, I'll have the same metadata entry match multiple samplesheet entries.

