---
title: "02-locus-HWE-evaluation"
author: "Diana Baetscher"
date: "2023-03-30"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



Using the complete dataset to evaluate HWE across loci and populations in the baseline.



Updates: use the NAs explicit genos file. 
Don't filter loci based on missing data at this point. 
Just look at reference pops for evaluating HWE.


Evaluating the loci that we're using for self-assignment and bycatch assignment.
These were originally derived from lcWGS data and the microhaplotype results suggest that the real variation present is potentially quite different.
Initially, I should probably just focus on the baseline genotypes rather than including everything.


To evaluate the markers, I need to subset the reference baseline genotypes for just the minimal amount of missing data.

```{r load-libraries}
library(tidyverse)
library(adegenet)
library(radiator)
library(DescTools)

```


```{r}
# read in rds file with genotypes
genos_long <- read_rds("../data/processed/called_genos_na_explicit.rds")

# just select the reference samples based on metadata
samplesheets <- read_rds("../data/processed/metadata_bycatch_and_reference_20230330.rds")

reference_samples <- samplesheets %>%
  filter(Sample_Plate %in% c("BFAL001", "BFAL002") &
           !is.na(Location),
         !`Loc-Abbr` %in% c("Oahu", "Kauai, Kilauea")) %>%
  arrange(sampleID) %>%
  select(sampleID, gtseq_run, id, Location, `Loc-Abbr`)

# combine the genotypes with the reference
ref_genos <- reference_samples %>%
  left_join(., genos_long, by = c("gtseq_run", "id")) 

```

Deal explicitly with duplicate samples:

## Some initial filters

### Take highest read-depth call for multiply-genotyped DNA_IDs

I'm not sure if there are any of these, but best to leave it in here...

Now, here is a harder operation: if an individual is multiply-genotyped, take the
genotype with the highest total read depth.  
```{r take-just-one}
# slow-ish function to get the total read depth column
tdepth <- function(a, d) {
  if(any(is.na(a))) {
    return(NA)
  }
  if(a[1]==a[2]) {
    return(d[1])
  } else {
    return(d[1] + d[2])
  }
  
}
# this takes the highest read-depth instance of each duplicately-genotyped individual.
geno_one_each <- ref_genos %>%
  group_by(sampleID, locus, gtseq_run) %>%
  mutate(total_depth = tdepth(allele, depth)) %>%
  ungroup() %>%
  arrange(sampleID, locus, total_depth, gtseq_run, depth) %>%
  group_by(sampleID, locus) %>%
  mutate(rank = 1:n()) %>% # this ranks all alleles for a given individual by depth
  ungroup() %>%
  filter(rank <= 2) # this takes the top ranked alleles and should remove duplicates
  

# how many samples now?
geno_one_each %>%
  filter(!is.na(sampleID)) %>%
  group_by(sampleID) %>% 
  select(sampleID, gtseq_run, id) %>%
  unique() %>%
  tally() 

```



Take a look at missing data overall:
```{r}
geno_one_each %>%
  ggplot(aes(x = reorder(locus, depth), y = reorder(sampleID, depth), fill = log10(depth))) +
  geom_tile()

```


Deal with missing data:


## Missing data in loci

How many loci and how many alleles?
```{r}
# alleles
geno_one_each %>%
  filter(!is.na(allele)) %>%
  select(locus, allele) %>%
  unique()

# loci
geno_one_each %>%
  filter(!is.na(allele)) %>%
  select(locus, allele) %>%
  unique() %>%
  group_by(locus) %>%
  tally() %>%
  arrange(desc(n))

```


In the reference, 458 alleles across 189 loci with 1-6 alleles per locus.

Some individuals are missing all their data, and some loci are missing all their data.
Beginning with loci:

```{r}
# missing data across loci
locs_to_toss <- geno_one_each %>%
  group_by(locus) %>%
  mutate(missingness = ifelse(is.na(allele), 1, 0)) %>%
  summarise(sum(missingness)) %>% # 189 loci x2 = total
  filter(`sum(missingness)`>151) %>% # more than 50% missing data (given 151 individuals)
  select(locus) # drop those loci for now and see how the assignment goes

# just the keepers
genos_locs_filtered <- geno_one_each %>%
  anti_join(., locs_to_toss)


```


Remove these 4 loci because of too much missing data in the reference samples.


## Missing data in individuals

Total number of loci = 185

Total number of samples = 151

```{r}
inds_to_toss <- genos_locs_filtered %>%
  group_by(sampleID) %>%
  mutate(missingness = ifelse(is.na(allele), 1, 0)) %>%
  summarise(sum(missingness)) %>%
  arrange(desc(`sum(missingness)`)) %>%
  #filter(`sum(missingness)` > 47) # remove samples with >25% missing data
 filter(`sum(missingness)` > 19) # remove samples with >10% missing data

# just the keepers
genos_locs_ind_filtered <- genos_locs_filtered %>%
  anti_join(., inds_to_toss)

```
23 samples removed with >10% missing data.



Now look at missing data:
```{r}
genos_locs_ind_filtered %>%
  ggplot(aes(x = reorder(locus, -depth), y = reorder(sampleID, -depth), fill = log10(depth))) +
  geom_tile()

```

Much better. Now move forward from here.


## Prep data for analysis


```{r}
# first make integers of the alleles
alle_idxs <- genos_locs_ind_filtered %>% 
  filter(!is.na(sampleID)) %>%
  dplyr::select(sampleID, locus, gene_copy, allele) %>%
  group_by(locus) %>%
  mutate(alleidx = as.integer(factor(allele, levels = unique(allele)))) %>%
  ungroup() %>%
  arrange(sampleID, locus, alleidx) # rubias can handle NA's, so no need to change them to 0's

# reformat
reference <- alle_idxs %>%
  inner_join(., reference_samples) %>%
  select(-allele, -gtseq_run, -id) %>%
  select(`Loc-Abbr`, sampleID, everything()) %>%
  rename(collection = `Loc-Abbr`, indiv = sampleID)


```


I need to finagle the dataset into a genind object.

```{r}
# read in data
reference_genos <- reference

# number of missing alleles (max = 185*2)
inds_to_toss_missing_data <- reference_genos %>%
  group_by(indiv) %>%
  filter(is.na(alleidx)) %>%
  group_by(indiv, collection) %>%
  tally() %>%
  arrange(desc(n)) %>%
  filter(n > 9)
  
# if I removed all the indivs with > 10 missing alleles, that's only 5 indivs.
# what would the distribution of indivs per population be? 
# still ok, I'm pretty sure.
reference_genos %>%
  anti_join(., inds_to_toss_missing_data) %>%
  group_by(collection, indiv) %>%
  tally() %>%
  tally()

```

I feel fine about that. Let's see if that makes any of the HWE calculations easier because less missing data?

```{r}
# SKIP THIS FOR NOW.
# ref_genos_low_missing_data <- reference_genos %>%
#   anti_join(., inds_to_toss_missing_data)

ref_genos_low_missing_data <- reference_genos

```



Make the df match the requirements for tidy_genomic_data
```{r}
long_df <- ref_genos_low_missing_data  %>%
  select(-gene_copy) %>%
  select(collection, everything()) %>%
  rename(INDIVIDUALS = indiv, STRATA = collection, MARKERS = locus, GT = alleidx)

```

Genotypes should be coded with 3 integers for each alleles. 6 integers in total for the genotypes. e.g. 001002 or 111333 (for heterozygote individual). 6 integers WITH separator: e.g. 001/002 or 111/333 (for heterozygote individual). The separator can be any of these: "/", ":", "_", "-", ".", and will be removed.


```{r}
# create 3 digit integers from the genotypes
long_df$GT3 <- Format(long_df$GT, ldigits = 3, digits = 0)

# fix NAs
long_df0s <- long_df %>%
  mutate(GT3 = ifelse(is.na(GT3), "000", GT3)) # I don't love that this creates potential artifacts!
```

Now combine the GT3 column per indiv/marker:
```{r}
# make the genos characters and then try pasting them as strings
long_df0s$GT3 <- as.character(long_df0s$GT3)

long_df3digit <- long_df0s %>%
  group_by(INDIVIDUALS, MARKERS) %>% 
  arrange(GT3, .by_group = TRUE) %>% 
  summarise(GENOTYPE = toString(GT3))
  
# paste strings together
long_df3digit$GENOTYPE <- gsub(", ","",long_df3digit$GENOTYPE)


# add back on species identity as strata
df_for_conversion <- long_df0s %>% 
  select(-GT, -GT3) %>%
  left_join(., long_df3digit) %>%
  unique() %>%
  rename(GT = GENOTYPE) %>%
  mutate(GT = ifelse(GT == "000000", NA, GT)) %>%
  filter(!str_detect(STRATA, "Kauai"))
  

df_for_conversion$STRATA <- as.factor(df_for_conversion$STRATA)

```

Double check - how many samples per population?
```{r}
df_for_conversion %>%
  select(INDIVIDUALS, STRATA) %>%
  unique() %>%
  group_by(STRATA) %>%
  tally()

```

Let's remove Lehua and Kilauea at this point because of so few samples.

```{r}
convert_df_wo_2pops <- df_for_conversion 
```



```{r convert-df-to-genind}
# use the radiator package for this conversion
genind_df <- write_genind(convert_df_wo_2pops)

```
Basic population genetic evaluation of the markers, following, in part, this:
https://bookdown.org/hhwagner1/LandGenCourse_book/WE_3.html


```{r}
sum_output <- summary(genind_df)

names(sum_output)

expected_hz <- as.data.frame(sum_output$Hexp)
observed_hz <- as.data.frame(sum_output$Hobs)


```

Expected heterozygosity (here: Hexp) is the heterozygosity expected in a population under HWE, and observed heterozygosity (here: Hobs) is the observed number of heterozygotes at a locus divided by the total number of genotyped individuals. Here are the global values (pooled across all populations):

```{r}
# expected heterozygosity per population
adegenet::Hs(genind2genpop(genind_df))
```


```{r}
Hobs <- t(sapply(seppop(genind_df), function(ls) summary(ls)$Hobs))
  Hexp <- t(sapply(seppop(genind_df), function(ls) summary(ls)$Hexp))
  {cat("Expected heterozygosity (Hexp):", "\n")
  round(Hexp, 2)
  cat("\n", "Observed heterozygosity (Hobs):", "\n")
  round(Hobs, 2)}
  
```

```{r}
par(mar=c(5.5, 4.5, 1, 1))
  Hobs.pop <- apply(Hobs, MARGIN = 1, FUN = mean)
  Hexp.pop <- apply(Hexp, 1, mean) 
  barplot(Hexp.pop, ylim=c(0,1), las=3, ylab="Expected heterozygosity")
  barplot(Hobs.pop, ylim=c(0,1), las=3, ylab="Observed heterozygosity")
```



```{r}
bfal_diversity <- data.frame(Pop = names(Hobs.pop),
                              n_samples = sum_output$n.by.pop,
                              Hobs = Hobs.pop,
                              Hexp = Hexp.pop)
                              #Ar = Richness$mean.richness)
as.tibble(bfal_diversity) %>%
  rename(Population = Pop, H_obs = Hobs, H_exp = Hexp) #%>%
  #write_csv("csv_outputs/reference_pop_hz_summary.csv")

```

chi^2: value of the classical chi-squared test statistic
df: degrees of freedom of the chi-squared test
Pr(chi^2 >): p-value of the chi-squared test (‘>’ indicates that the alternative is ‘greater,’ which is always the case for a chi-squared test)
Pr.exact: p-value from an exact test based on Monte Carlo permutation of alleles (for diploids only). The default is B = 1000 permutations (set B = 0 to skip this test). Here we use the function ‘round’ with argument ‘digits = 3’ to round all values to 3 decimals.

https://rdrr.io/cran/pegas/man/hw.test.html

In this case, the degrees of freedom depend on the number alleles for a given locus.

```{r}
# HWE test for all loci - but really what we want is all loci in each population (farther below)
hwe_output <- round(pegas::hw.test(genind_df, B = 1000), digits = 3)

# which ones are statistically sign.
rownames_to_column(data.frame(hwe_output)) %>%
  filter(Pr.exact < 0.01)

```

28 loci that are out of HWE globally.

After looking at the loci x population HWE info, I can remove any that are out of HWE in the majority of populations.


```{r}
# Chi-squared test: p-value
HWE.test <- data.frame(sapply(seppop(genind_df), 
                              function(ls) pegas::hw.test(ls, B=0)[,3]))
HWE.test.chisq <- t(data.matrix(HWE.test))
{cat("Chi-squared test (p-values):", "\n")
round(HWE.test.chisq,3)}

# these are the p-values for the per-pop calcs
loci_to_toss_hwe <- rownames_to_column(as.data.frame(HWE.test.chisq), var = "population") %>%
  pivot_longer(cols= 2:length(rownames_to_column(as.data.frame(HWE.test.chisq))), names_to = "locus", values_to = "p_val") %>%
  filter(p_val < 0.05) %>% # which p-values are significant?
  group_by( locus) %>%
  tally() %>% # how many loci are out of HWE in how many populations?
  arrange(desc(n)) %>%
  filter(n>3)

```

Remove loci that are statistically out of HWE in >3 populations (the majority).


That's only 4 loci.

```{r}
loci_to_toss_hwe %>%
  write_csv("csv_outputs/loci_to_toss_hwe.csv")

loci_to_keep <- rownames_to_column(as.data.frame(HWE.test.chisq), var = "population") %>%
  pivot_longer(cols= 2:length(rownames_to_column(as.data.frame(HWE.test.chisq))), names_to = "locus", values_to = "p_val") %>%
  select(locus) %>%
  unique() %>%
  anti_join(., loci_to_toss_hwe)
  


loci_to_keep %>%
  write_csv("csv_outputs/loci_to_keep_hwe.csv")
# from here, I can generate a "final" dataset.
```