---
title: "manuscript-figures"
author: "diana baetscher"
date: "2023-12-06"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Tightening up some figures for the BFAL manuscript.
```{r load-libraries}
library(tidyverse)
library(readxl)

```


Reading in the data from Jessie:
```{r}
#dat <- read_csv("../data/BFAL_spatial_UDs_4.25.23.csv")

final_output <- read_csv("../data/BFAL_GSI_bands_output_20231208_latlong.csv") %>%
    filter(CollYear > 2009) # 28 birds pre-2010

final_output %>%
  group_by(FisheriesName, colony) %>%
  tally()

# remove those pre-2010 and then move the dataset forward with just samples with spatial information
bfal <- final_output %>%
  filter(!is.na(lat)) %>%
  mutate(colony = str_replace(colony, "NWHI, ", "")) %>%
  filter(colony %in% c("Midway", "Kure", "FFS")) #%>%
    # group_by(colony) %>%
    # tally()
  
# 617 birds with spatial information
  

banded_birds_no_loc_keepers <- final_output %>%
  filter(`BandY/N` == T & is.na(Location) & PofZ > 0.9 & z_score > -3)

bands_to_toss <- final_output %>%
  filter(`BandY/N` == T & is.na(Location)) %>%
  anti_join(., banded_birds_no_loc_keepers)

# bfal <- final_output %>%
#   anti_join(., bands_to_toss) %>%
#   inner_join(., dat) # for now this is an inner join. Hopefully there's more spatial data for the 656 samples that are in the final dataset
# 
# bfal
```
7 banded bycatch samples without location information and without confident GSI assignments. Remove these from the dataset.


```{r color-palette}
# set colors
mypalette <- c("dodgerblue", "#006d2c", "skyblue", "darkslateblue", "#74c476",  # Hawaii - greens, blues
               "tomato", "#31a354") # Japan - red
               

# Kure = dodgerblue
# Midway = darkslateblue
# FFS = "#74c476"
```
```{r population-estimates}
pop_est <- read_csv("../data/bfal_pop_estimates.csv")

prop_total_pop <- pop_est %>%
  filter(colony != "Total") %>%
  mutate(total_pop = sum(population_estimate)) %>%
  mutate(perc_of_total = population_estimate/total_pop) %>%
  mutate(GenCollArea = "Total population") %>%
  mutate(colony = colony_abbr) # fix this once we figure out what's needed for Lehua


```


Make barplot based on population proportions
```{r}
final_bycatch_prop <- final_output %>%
  anti_join(., bands_to_toss) %>%
  group_by(colony, GenCollArea) %>%
  mutate(colony = ifelse(colony == "NWHI, FFS", "FFS", colony)) %>%
  mutate(colony = ifelse(colony == "NWHI, Kure", "Kure", colony)) %>%
  mutate(colony = ifelse(colony == "NWHI, Midway", "Midway", colony)) %>%
  mutate(colony = ifelse(colony == "NWHI, Laysan", "Laysan", colony)) %>%
  mutate(colony = ifelse(colony %in% c("Kauai, Kilauea", "Kauai, Lehua"), "Other", colony)) %>%
  mutate(colony = ifelse(is.na(colony), "Unassigned", colony)) %>%
  tally() %>%
  arrange(GenCollArea) %>%
  ungroup() %>%
  group_by(GenCollArea) %>%
  mutate(area_bycatch = sum(n)) %>%
  group_by(colony, GenCollArea) %>%
  mutate(col_bycatch_region = sum(n)) %>%
  mutate(col_bycatch_region_prop = col_bycatch_region/area_bycatch) %>%
  # calculate the bycatch proportions for all fisheries combined
  ungroup() %>%
  mutate(total_bycatch = sum(n)) %>%
  group_by(colony) %>%
  mutate(colony_total = sum(n)) %>%
  mutate(perc_of_total = colony_total/total_bycatch) %>%
  ungroup() 
  
all_fish_data <- final_bycatch_prop %>%
  dplyr::select(colony, perc_of_total) %>%
  mutate(GenCollArea = "All fisheries") %>%
  unique()

# add in proportion of total population
prop_total_pop %>%
  left_join(., final_bycatch_prop, by = c("colony_abbr" = "colony")) 


combined_prop_df <- final_bycatch_prop %>%
  bind_rows(., prop_total_pop) %>%
  bind_rows(all_fish_data) %>%
  mutate(col_bycatch_region_prop = ifelse(is.na(col_bycatch_region_prop), perc_of_total, col_bycatch_region_prop)) %>%
  filter(!is.na(colony),
         GenCollArea != "WA") %>%
  mutate(GenCollArea = ifelse(GenCollArea == "AK", "Alaska fisheries", GenCollArea)) %>%
    mutate(GenCollArea = ifelse(GenCollArea == "HI", "Hawaii fisheries", GenCollArea))

  
```
Remove WA (low sample sizes)
Add all fisheries combined (including WA)

```{r barplot-with-proportions}
area_order <- combined_prop_df$GenCollArea <- factor(combined_prop_df$GenCollArea, levels=c("Total population", "All fisheries", "Alaska fisheries", "Hawaii fisheries"))

bycatchA <- combined_prop_df %>%
  mutate(colony = ifelse(colony == "FFS", "French Frigate Shoals", colony)) %>%
  ggplot(aes(x = GenCollArea, y = col_bycatch_region_prop, fill = colony)) +
  geom_bar(stat = "identity", position = "stack") +
  #facet_grid(rows = vars(GenCollArea)) +
  theme_bw() +
  labs(
    fill = "BFAL breeding \ncolony",
    y = "Proportion",
    x = " "
  ) +
  scale_y_continuous(limits = c(0, 1), expand = c(0,0)) +
  scale_fill_manual(values = c("#74c476", "dodgerblue", "#006d2c", "darkslateblue", "skyblue", "tomato", "gray")) +
  theme(
    axis.title.x = element_text(margin = margin(t=10)),
      axis.title.y = element_text(margin = margin(r = 10))

  )

bycatchA

ggsave("pdf_outputs/colony_bycatch_barplot.png", width = 7, height = 4)
```
Alternative plot with differences between the population and the fisheries:

```{r difference-plot-for-comparison}
# because I had to futz with the dataframe earlier, there are odds and ends out of place
bycatch_diff <- combined_prop_df %>%
  mutate(colony = ifelse(colony == "FFS", "French Frigate Shoals", colony)) %>%
  dplyr::select(colony, GenCollArea, col_bycatch_region_prop) %>%
  pivot_wider(names_from = "GenCollArea", values_from = col_bycatch_region_prop) %>%
  pivot_longer(cols = c(`All fisheries`, `Alaska fisheries`, `Hawaii fisheries`), names_to = "fishery", values_to = "prop_bycatch") %>%
  mutate(perc_diff = (prop_bycatch-`Total population`)*100) %>%
  filter(!is.na(perc_diff)) 
# reorder the factor
bycatch_diff$fishery <- factor(bycatch_diff$fishery, levels= c("All fisheries", "Alaska fisheries", "Hawaii fisheries"))
bycatch_diff$colony <- factor(bycatch_diff$colony, levels = c("French Frigate Shoals", "Kure", "Laysan", "Midway", "Torishima", 
                                                              "Other"))

bycatchB <- bycatch_diff %>%
  ggplot(aes(x = fishery, y = perc_diff, color = colony)) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, col = "black", linetype = "dotted") +
   theme_bw() +
  labs(
    color = "Bycatch \ncolony of origin",
    y = "% diff. bycatch-population prop.",
    x = " "
  ) +
  #scale_y_continuous(limits = c(0, 1), expand = c(0,0)) +
  scale_color_manual(values = c("#74c476", "dodgerblue", "#006d2c", "darkslateblue", "tomato", "skyblue")) +
  theme(
    axis.title.x = element_text(margin = margin(t=10)),
      axis.title.y = element_text(margin = margin(r = 10))

  )
  
bycatchB

ggsave("pdf_outputs/colony_bycatch_diff_plot.png", width = 6, height = 3.5)

```

```{r combine-bycatch-plots}
library(patchwork)

bycatchA + bycatchB + plot_annotation(tag_levels = "A") + plot_layout(ncol = 1)

ggsave("pdf_outputs/colony_bycatch_diff_plot_2panel.png", width = 7, height = 6)
```




## Spatial analysis with utilization distributions

Copying over Jessie's code and modifying, then modifying more with Jon's input:

```{r spatial-analysis-maps}
#load packages (could probably remove some of these - calling in a bunch just in case)
library(tidyverse)
library(adehabitatHR)
library(ggplot2)
library(magrittr)
library(sp)
library(sf)
library(raster)
library(rgdal)
library(mapdata)
library(dplyr)

bfal$colony <- factor(bfal$colony, levels = c("FFS", "Kure", "Midway"), 
                  labels = c("FFS (486)", "Kure (46)", "Midway (57)"))

bfal$long360=ifelse(bfal$long < 0,bfal$long + 360, bfal$long)

#colony locations
colony_locs <- read_csv("../data/latlon_df.csv")


```



```{r spatial-analysis-maps-part2}
#summarize to double check
bfal %>% group_by(colony) %>% 
  tally()

# there are 617 birds at this stage, AK, HI, and WA combined
# only includes birds with band data or GSI with PofZ >=.9, and must have spatial data

# test plot
plot(bfal$long360, bfal$lat, asp = 1)

bfal %>%
  dplyr::select(sampleID, lat, long, long360)

points = st_as_sf(bfal, coords = c("long360", "lat"), crs = 4326)
plot(st_geometry(points), pch=1, col="navy")

# make kernels

# Make a spatial points data.frame (SPDF)
spdf <- SpatialPointsDataFrame(coords=cbind(bfal$long,
  # bfal$long360,
                                            bfal$lat),
                               data=data.frame(id=bfal$colony),
                               proj4string = CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +wrap_lon=180"))

# Project data into laea
spdf.t <- spTransform(spdf,
                      # CRS("+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154")
                      CRS("+proj=aea +lat_1=30 +lat_2=70 +lat_0=52 +lon_0=-170 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs") #https://spatialreference.org/ref/sr-org/north-pacific-albers-conic-equal-area/
                      ) ## this projection might be wrong - check with Jon?

# calculate kernelUD
#jb try to up h factor and see what happens (more smoothing) ## get more details from Jon on the smoothing factor
ud <- kernelUD(spdf.t, 
               h = 100000,
               # h="href",
               # h=200000, #200km; 100km as used previously seemed a bit small, but if you want more detail and independent ud50 blobs maybe go back to 100km
               grid=200, #
               same4all=TRUE #the default for this is FALSE but I think you might want to calc the UDs for each ID (colony) over the same extent
               )
ud90 <- getverticeshr(ud, percent=90, standardize=T)
ud50 <- getverticeshr(ud, percent=50, standardize=T)

ud50sf<-st_as_sf(ud50) #I am just keeping these separate
ud90sf<-st_as_sf(ud90)

world <-st_crop(st_transform(st_as_sf(maps::map("world", plot = FALSE, fill = TRUE)), 
                             st_crs(ud90sf)),
                             1.15*st_bbox(ud90sf)) %>% #crop to just beyond extent of ud90
  filter(ID!="Antarctica")

p3 <-ggplot() + 
  geom_sf(data=ud90sf, aes(fill=id), alpha=0.2) +
  geom_sf(data=ud50sf, aes(fill=id), alpha=0.8) + 
  geom_sf(data=world, fill="gray30",color=NA) +
  facet_wrap(~id) +
  labs(fill = "UD by \ncolony",
       x = "",
       y = "")+
  theme_bw() +
  scale_x_continuous(breaks = c(-175, -155, -135), labels = c("175˚W", "155˚W", "135˚W"), expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("#74c476", "dodgerblue", "slateblue")) +
  theme(
    axis.text.x = element_text(size = 7),
    axis.text.y = element_text(size = 7)
  ) 

p3 

# ann_text <- data.frame(x = -165, y = 20, lab = "90% UD",
#                        id = factor(`FFS (486)`, levels = c("FFS (486)", "Kure (46)", "Midway (57)")))

# p + geom_text(data = ann_text,label = "Text")

ggsave(plot=p3,"pdf_outputs/BFAL_UDs_3panels.png", width = 8, height = 4)

```

